\morus: Biases in the Stream

### 2018-05-10 morusAC_00_Abstract.tex
minor rewrites in abstract and contributions

\begin{abstract}
  \morus is a high-performance authenticated encryption algorithm
  submitted to the CAESAR competition, and recently selected as a
  finalist.  There are three versions of \morus: \morus[640] with a
  640-bit state and a 128-bit key, and \morus[1280] with a 1280-bit
  state and a 128-bit or 256-bit key; for all versions the security
  claim for confidentiality matches the key size.  In this paper, we
  analyze the components of this algorithm (initialization, state
  update and tag generation), and report several results.

  Our main result is on the full state update of \morus: we found a
  linear bias in the keystream that can be used to distinguish the
  output from random, and to recover some plaintext bits in the
  broadcast setting.  For \morus[1280], the bias is $2^{-76}$,
  which can be identified after around $2^{152}$ encryptions, less than
  would be expected for a 256-bit secure cipher.  On the other hand, for
  \morus[640], the bias is $2^{-73}$, which does not threaten
  the security of the cipher.

% More precisely, we show that a linear correlation exists between some bits of the plaintext and ciphertext. Moreover the bias depends purely on the plaintext, and not on the secret key nor the nonce, hence restricting the single use of nonce or even rekeying does not prevent the attacks.
%
% From the nature of the linear cryptanalysis, the attack works in the known plaintext attacks. Moreover, if a fixed plaintext is encrypted multiple times, the correlation can be used to recover a 1-bit information of the plaintext. The attack principally has larger impact because it can be used to recover more unknown bits of the plaintext, provided an initial segment of the plaintext is known. 
%
  To identify the bias, we exploit rotational symmetries in \morus
  using linear masks that are invariant by word-rotations of the
  state.  This motivates us to introduce word-oriented versions of
  \morus called \minimorus, which help the
  analysis. The attack has been implemented and verified on
  \minimorus, where it yields a bias of $2^{-16}$.

@@

  We also study reduced versions of the initialization and
  finalization of \morus to evaluate the security margin of these
  components.  We show a forgery attack when the finalization is
  reduced from 10 rounds to 3, and a key-recovery attack in the
  nonce-misue setting when the initialization is reduced from 16 to 10
  steps.  These additional results do not threaten full \morus, but
  studying all aspects of the design is useful to understand its
  strengths and weaknesses.

>>>>

\begin{abstract}
  \MORUS is a high-performance authenticated encryption algorithm
  submitted to the CAESAR competition, and recently selected as a
  finalist.  There are three versions of \MORUS: \MORUS[640] with
  %a 640-bit state and
  a 128-bit key, and \MORUS[1280] with
  % a 1280-bit state and
  a 128-bit or 256-bit key. For all versions the security
  claim for confidentiality matches the key size.  In this paper, we
  analyze the components of this algorithm (initialization, state
  update and tag generation), and report several results.

  As our main result, we exhibit a
  linear bias in the keystream of full \MORUS, which can be used to distinguish its output
  from random, and to recover some plaintext bits in the
  broadcast setting.  For \MORUS[1280], the bias is $2^{-76}$,
  which can be exploited after around $2^{152}$ encryptions, less than
  would be expected for a 256-bit secure cipher.  For
  \MORUS[640], the same attack results in a bias of $2^{-73}$, which does not threaten
  the security of the cipher.

% More precisely, we show that a linear correlation exists between some bits of the plaintext and ciphertext. Moreover the bias depends purely on the plaintext, and not on the secret key nor the nonce, hence restricting the single use of nonce or even rekeying does not prevent the attacks.
%
% From the nature of the linear cryptanalysis, the attack works in the known plaintext attacks. Moreover, if a fixed plaintext is encrypted multiple times, the correlation can be used to recover a 1-bit information of the plaintext. The attack principally has larger impact because it can be used to recover more unknown bits of the plaintext, provided an initial segment of the plaintext is known. 
%
  To identify this bias, we make use of rotational symmetries in \MORUS
  using linear masks that are invariant by word-rotations of the
  state.  This motivates us to introduce single-word versions of
  \MORUS called \MiniMORUS, which aid in the
  analysis. The attack has been implemented and verified on
  \MiniMORUS, where it yields a bias of $2^{-16}$.

@@

  We also study reduced versions of the initialization and
  finalization of \MORUS, aiming to evaluate the security margin of these
  components.  We show a forgery attack when finalization is
  reduced from 10 rounds to 3, and a key-recovery attack in the
  nonce-misue setting when initialization is reduced from 16 to 10
  steps.  These additional results do not threaten full \MORUS, but
  studying all aspects of the design is useful to understand its
  strengths and weaknesses.

### 2018-05-10 morusAC_01_Introduction.tex
minor rewrites in abstract and contributions

\subsubsection*{Related Work.}
In the \MORUS submission document, the designers discuss the security of \MORUS against
several attacks, including algebraic, differential, and guess-and-determine attacks.
The main focus is on differential properties, and not many details are given for other attack vectors.
In third-party analysis,
Mileva et~al.~\cite{balkancryptsecMilevaDV15} propose a distinguisher in the nonce-reuse setting and practically evaluate the differential behaviour of toy variants of \MORUS.
Dwivedi et~al.~\cite{cryptoeprint:2016:1053} discuss the applicability of SAT solvers for state recovery, but the resulting complexity of $2^{370}$ for \MORUS[640] is well beyond the security claim.
Dwivedi et~al.~\cite{secryptDwivediMW17} also propose key-recovery attacks for \MORUS[1280] if initialization is reduced to 3.6 out of 16 steps, and discuss the security of \MORUS against internal differentials and rotational cryptanalysis.
Salam et~al.~\cite{trustcomSalamSBDPW17} apply cube attacks to obtain distinguishers for up to 5 out of 16 steps of the initialization of \MORUS[1280] with negligible complexity.
Additionally, 
Kales et~al.~\cite{cryptoeprint:2017:1137} and
Vaudenay and Viz\'{a}r \cite{cryptoeprint:2017:1147}
independently propose state-recovery and forgery attacks on \MORUS in a nonce-misuse setting with negligible data and time complexities.
Finally, a keystream bias of a similar nature as our main attack was uncovered by Minaud \cite{sacryptMinaud14} on the authenticated cipher \cipher{AEGIS} \cite{AEGIS,sacryptWuP13}, another CAESAR finalist which shares the same overall structure as \MORUS, but uses a very different, \cipher{AES}-based state update function.

\subsubsection*{Our Contributions.}
We analyze the security of \MORUS and present a key\-stream distinguisher based on
linear approximations for full \MORUS[1280].
In addition, we provide results for round-reduced \MORUS with reduced initialization or finalization.

Our main result is a linear approximation \cite{eurocryptMatsui93,eurocryptMatsuiY92} linking ciphertext bits of five consecutive encryption blocks.
%As a consequence of this trail, we exhibit a biased linear combination of ciphertext bits, which holds across any given sequence of five consecutive encryption steps.
%Technically, this assumes that the inner state of the cipher at the start of the trail is uniformly random, but this is a fair assumption since the inner state should be indistinguishable from uniform randomness by design.
%Moreover, that is the only assumption required; in particular, the secret key or the nonce used to initialize the inner state of MORUS are irrelevant: the bias is independent of their value.
More precisely, we show that when encrypting a fixed plaintext block multiple times, a linear correlation exists between some output bits of the cipher.
Moreover, the bias depends purely on the plaintext, and not on the secret key of the cipher.
In principle, this property could be used to recover unknown bits of a plaintext encrypted a large number of times, provided an initial segment of the plaintext is known.
For \MORUS[1280] with 256-bit keys, the linear bias is $2^{-73}$
\todo{double-check bias vs. correlation everywhere}
and can be exploited using about $2^{152}$ encrypted blocks.
\todo{double-check}
We note that our attack is nonce-respecting, and that rekeying does not prevent it: the biases are independent of the secret encryption key and nonce, and can be exploited as long as a given plaintext block is encrypted sufficiently often, regardless of whether each encryption uses a different key.

To obtain this result, we propose a simplified abstraction of \MORUS, called \MiniMORUS, which takes advantage of certain rotational symmetries in \MORUS.
We observe that the actual bias of the approximation is better than suggested by the construction of the characteristics.
To confirm the validity of our analysis,
we practically verified the bias of the linear approximation for \MiniMORUS and of partial approximations for \MORUS.
In addition, we symbolically evaluated the relevant equations.

In addition to this attack on full \MORUS, we also analyze the security of \MORUS with round-reduced initialization (in a nonce-misuse setting) or finalization.
While this analysis does not threaten full \MORUS, it complements the main result to provide a better understanding of the \MORUS design approach.
We present a forgery attack for round-reduced \MORUS[1280] with success probability $2^{-88}$ for a 128-bit tag if the finalization is reduced to 3 out of 10 steps. This nonce-respecting attack is based on a differential analysis of the padding rule.
The second result targets a round-reduced initialization with 10 out of 16 steps, and extends a state-recovery attack (e.g., in a nonce-misuse setting) into a key-recovery attack.

>>>>

\subsubsection*{Related Work.}
In the \MORUS submission document, the designers discuss the security of \MORUS against
several attacks, including algebraic, differential, and guess-and-determine attacks.
The main focus is on differential properties, and not many details are given for other attack vectors.
In third-party analysis,
Mileva et~al.~\cite{balkancryptsecMilevaDV15} propose a distinguisher in the nonce-reuse setting and practically evaluate the differential behaviour of toy variants of \MORUS.
Dwivedi et~al.~\cite{cryptoeprint:2016:1053} discuss the applicability of SAT solvers for state recovery, but the resulting complexity of $2^{370}$ for \MORUS[640] is well beyond the security claim.
Dwivedi et~al.~\cite{secryptDwivediMW17} also propose key-recovery attacks for \MORUS[1280] if initialization is reduced to 3.6 out of 16 steps, and discuss the security of \MORUS against internal differentials and rotational cryptanalysis.
Salam et~al.~\cite{trustcomSalamSBDPW17} apply cube attacks to obtain distinguishers for up to 5 out of 16 steps of the initialization of \MORUS[1280] with negligible complexity.
Additionally, 
Kales et~al.~\cite{cryptoeprint:2017:1137} and
Vaudenay and Viz\'{a}r \cite{cryptoeprint:2017:1147}
independently propose state-recovery and forgery attacks on \MORUS in a nonce-misuse setting with negligible data and time complexities.
Finally, a keystream bias similar in nature to our main attack was uncovered by Minaud \cite{sacryptMinaud14} on the authenticated cipher \cipher{AEGIS} \cite{AEGIS,sacryptWuP13}, another CAESAR finalist, which shares the same overall structure as \MORUS, but uses a very different, \cipher{AES}-based state update function.

\subsubsection*{Our Contributions.}
Our main contribution is a keystream distinguisher on full \MORUS[1280], built from linear approximations of its core \StateUpdate{} function.
In addition, we provide results for round-reduced \MORUS, targeting both the initialization or finalization phases of the cipher.

In more detail, our main result is a linear approximation \cite{eurocryptMatsui93,eurocryptMatsuiY92} linking ciphertext bits across five consecutive encryption blocks.
%As a consequence of this trail, we exhibit a biased linear combination of ciphertext bits, which holds across any given sequence of five consecutive encryption steps.
%Technically, this assumes that the inner state of the cipher at the start of the trail is uniformly random, but this is a fair assumption since the inner state should be indistinguishable from uniform randomness by design.
%Moreover, that is the only assumption required; in particular, the secret key or the nonce used to initialize the inner state of MORUS are irrelevant: the bias is independent of their value.
As a consequence, when encrypting a fixed plaintext segment multiple times, a linear correlation exists between some output bits of the cipher.
Moreover, the bias depends purely on the plaintext, and not on the secret key of the cipher.
In principle, this property could be used to recover unknown bits of a plaintext encrypted a large number of times, provided an initial segment of the plaintext is known.
For \MORUS[1280] with 256-bit keys, the linear bias is $2^{-76}$
\todo{double-check bias vs. correlation everywhere}
and can be exploited using about $2^{152}$ encrypted blocks.
\todo{double-check}

To the best of our knowledge, this is the first attack on full \MORUS in the nonce-respecting setting.
We also note that rekeying does not prevent the attack: the biases are independent of the secret encryption key and nonce, and can be exploited as long as a given plaintext segment is encrypted sufficiently often, regardless of whether each encryption uses a different key.

To obtain this result, we propose a simplified abstraction of \MORUS, called \MiniMORUS. \MiniMORUS takes advantage of certain rotational symmetries in \MORUS and simplifies the description and analysis of the attack. We then show how the attack can be extended from \MiniMORUS to the real \MORUS.
%We observe that the actual bias of the approximation is better than suggested by the construction of the characteristics.
To confirm the validity of our analysis,
we practically verified the bias of the full linear trail for \MiniMORUS, as well as the bias of trail fragments for the full \MORUS.
Our analysis is also backed by a symbolic evaluation of the full trail equation and its bias on all variants of \MORUS.

In addition to the previous attack on full \MORUS, we provide two secondary results: (1) we analyze the security of \MORUS against forgery attacks with round-reduced finalization; and (2) we analyze its security against key recovery in a nonce-misuse setting, with round-reduced initialization.
While this extra analysis does not threaten full \MORUS, it complements the main result to provide a better overall understanding of the security of \MORUS.
More precisely, we present a forgery attack for round-reduced \MORUS[1280] with success probability $2^{-88}$ for a 128-bit tag if the finalization is reduced to 3 out of 10 steps. This nonce-respecting attack is based on a differential analysis of the padding rule.
The second result targets round-reduced initialization with 10 out of 16 steps, and extends a state-recovery attack (which can be mounted e.g. in a nonce-misuse setting) into a key-recovery attack.
